Temporal ordering and registration of images in studies of developmental dynamics

### 2014-09-01 image_analysis_paper2/SI_text.tex
added yannis comments

The original image resolution is $1024 \times 1024$ pixels for the fixed images, and $512 \times 512$ pixels for the live images; all images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
%
A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
%
When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
%
The same laser settings are maintained within each experiment.
%
The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
%
The Canny method is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
%
A constant $10$-pixel border is then added to each image. 
%
Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
%
The nuclear signal is blurred using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function).
%
For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals.

@@

We will demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
%
The relatively simple dynamics of this data set will allow us to easily visualize and illustrate the main features of the different algorithms.
%
Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

\begin{figure*}
\includegraphics{figS3}
\caption{Eigenvalue spectra. {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3. Note that there is a gap after the third coordinate. {\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3. Note that coordinates 2 and 3 are harmonics/functions of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional. {\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4. Note that there is a gap after the first coordinate; we can conclude that the data set is effectively one-dimensional. }
\end{figure*}

>>>>

The original image resolution is $1024 \times 1024$ pixels for the fixed images, and $512 \times 512$ pixels for the live images; all images are subsampled to $100 \times 100$ pixels for vector diffusion maps analysis.
%
A scale bar for a representative image is shown in \fig~\ref{fig:ap_dv}.
%
When imaging, the laser settings are calibrated so that images have good dynamic range, while not saturating the detector.
%
The same laser settings are maintained within each experiment.
%
The intensity of each channel is integer-valued, and ranges from $0$ (no signal) to $255$ (saturated). 
%
The Canny method \cite{canny1986computational} is used to detect the edges of the nuclear signal in each image (using the \texttt{edge} function in MATLAB\textsuperscript{\textregistered}), and images are cropped to these edges to remove any size effects.
%
A constant $10$-pixel-wide border is then added to each image. 
%
Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of the nuclear signal.
%
The nuclear signal is blurred using a disc filter with a $5$~pixel radius (using the \texttt{imfilter} function).
%
For the multichannel images, the nuclear signal is scaled by half (after normalization and blurring) relative to the other signals, so that the total nuclear signal across the entire image is of a comparable magnitude to the other signals.

@@

We demonstrate the algorithms for registration and temporal ordering using a synthetic data set.
%
The relatively simple dynamics of this data set allows us to easily visualize and illustrate the main features of the different algorithms.
%
Motivated by the geometry of our images, we construct a sequence of concentration profiles defined on a ring, and rotate each ring randomly around its center; an example is shown in \fig~\ref{subfig:1d_example}.
%
Rotation of the ring corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown at the bottom of \fig~\ref{subfig:1d_example} (the symmetry group is $SO(2)$, the group of all two-dimensional proper rotations).
%
Each concentration profile is a noisy Gaussian (shown in \fig~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as a function of ``time".
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors (the corresponding symmetry group for the discretized profiles is $\mathbb{Z}_{100}$, the group of integers modulo $100$).
%
\fig~\ref{subfig:1d_unaligned_unordered} shows the entire data set; the concentration profiles have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

@@

\begin{figure*}
\includegraphics{figS3}
\caption{Eigenvalue spectra. {\it (A)} Eigenvalue spectrum for the live imaging data set presented in \fig~3. Note that there is a gap after the third eigenvalue. {\it (B), (C)} Second and third embedding coordinates, respectively, versus the first embedding coordinate for the live imaging data set presented in \fig~3. Note that coordinates 2 and 3 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional. {\it (D)} Eigenvalue spectrum for the live imaging data set presented in \fig~4. Note that there is a gap after the first eigenvalue; we can conclude that the data set is effectively one-dimensional. }
\end{figure*}

### 2014-09-01 image_analysis_paper2/pnas_draft.tex
added yannis comments

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
In one of the most commonly used experimental formats, developmental progress is reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, such snapshots must be spatially registered and ordered in time.
%
Currently, registration and ordering of images is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the size of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We validate our approach using a live imaging data set where the correct rotational orientation and temporal order are known.
%
Finally, we demonstrate that our approach can successfully extract a representative developmental trajectory from a collection of images of fixed {\it Drosophila} embryos.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parametrization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
%
This idea of locality is denoted by the edges in \fig~\ref{subfig:dmaps1}.
%
Data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parametrization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images are known {\em a priori}.
%
This data set was obtained through live imaging of a {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
%
Often in practice, each images comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
%
The resulting data set is therefore spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

@@

\fig~\ref{subfig:fixed_images_registered_ordered} shows the images from \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
%
However, because of interembryo variability, the selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:fixed_images_average_trajectory}, serve as a summary of the stereotypic developmental dynamics.
%
We can now easily see the developmental progression, which is consistent with the known dynamics. 
%
dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

We can use prior knowledge about the developmental system to evaluate the quality of our registration and ordering. 
%
The Twi signal is known to form a single peak at the ventralmost point of the embryo. 
%
We found that the standard deviation in the location of this peak in the set of registered images was 8$^\circ$,
indicating that the algorithm successfully aligns the ventralmost points of the images. 
%
Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering. 
%
We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.  
%
The ranks from this expert-based ordering for the images in \fig~\ref{subfig:fixed_images_registered_ordered} are indicated, and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.

Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
%
In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $15$~seconds on an Intel Core i7 2.93 GHz processor. %for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
%
The computational time is a function of the number of images in the data set, the number of pixels in each point, and the number of angles used to compute the pairwise rotations (see \SI). 
%
Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the number of angles.
%
Furthermore, the computation of the pairwise rotational alignments, which is the most time-intensive portion of the calculation, is trivially parallelizable.
%
The requisite user intervention and parameter tuning required for our method is relatively minimal.
%
Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
%
Two algorithmic parameters, the number of angles to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
%
We found that the results are robust to both of these parameters, and that consistent parameter values were appropriate for all data sets presented here. 
%
%In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
%
The tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, especially when the data sets are large.  
%
Therefore, this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

We presented a unified approach to temporal ordering and registration of images in studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
In parallel, temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data are vectors describing the amounts of different chemical species.
%
Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
The presented approach addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace.
%; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
%
%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications. 
%
Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of developing {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other imaging data sets.

@@

Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
This necessitates the development of automated methodologies which are sufficiently general to address a wide variety of applications to organize such large data sets.
%
Vector diffusion maps allows us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with coordinates for each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several coordinates that serve as a good parameterization of the data set.
%
This parameterization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages. 
%
Images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.  
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

>>>>

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
In one of the most commonly used experimental formats, developmental progress is reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, such snapshots must be spatially registered and ordered in time.
%
Currently, registration and ordering of images is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We validate our approach using a live imaging data set where the correct rotational orientation and temporal order are known.
%
Finally, we demonstrate that our approach can successfully extract a representative developmental trajectory from a collection of images of fixed {\it Drosophila} embryos.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parametrization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parametrization  which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{subfig:dmaps1}:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parametrization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

To validate the proposed approach, we applied our algorithm to a data set where the true temporal order and rotational orientation of the images were known {\em a priori}.
%
This data set was obtained through live imaging of a {\it Drosophila} embryo during the twenty minutes spanning the late stages of cellularization through early gastrulation.
%
During this time window, the ventral furrow is formed, where the ventral side buckles towards the center of the embryo, internalizing the future muscle cells and forming a characteristic ``omega'' shape.
%
Germband extension then causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side \cite{leptin2005gastrulation}.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing a similar ``omega'' to appear on the dorsal side.

@@

We have analyzed how our algorithm performs on a model data set, where all images come from a single embryo. 
%
Often in practice, each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting an image from a random live imaging experiment at each time point.
%
The resulting data set is therefore spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.81, indicating that the algorithm can accurately recover the temporal order even under noisy conditions.

@@

\fig~\ref{subfig:fixed_images_registered_ordered} shows the selected images in \fig~\ref{subfig:fixed_images_unregistered_unordered}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set (see \ref{fig:full_fixed_data}).
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI). 
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:fixed_images_average_trajectory}, serve as a summary of the stereotypic developmental dynamics.
%
We can now easily see the developmental progression, which is consistent with the known dynamics: 
dpERK first appears as two lateral peaks at the ventral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{subfig:fixed_images_average_trajectory}.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).

We can use prior knowledge about the developmental system to evaluate the quality of our registration and ordering. 
%
The Twi signal is known to form a single peak at the ventralmost point of the embryo. 
%
We found that the standard deviation in the location of this peak in the set of registered images was 8$^\circ$,
indicating that the algorithm successfully aligns the ventralmost points of the images. 
%
Because the developmental time of each embryo cannot be easily estimated, we have few options for evaluating the quality of our temporal ordering. 
%
We compared the ordering obtained from vector diffusion maps to the ordering provided by a trained embryologist who is knowledgeable about the developmental progression and the important image features.  
%
The ranks from the ordering provided by the embryologist, which we will refer to as the ``expert rank'', are indicated for the images in \fig~\ref{subfig:fixed_images_registered_ordered} , and the rank correlation (see \fig~\ref{subfig:fixed_images_rank_corr}) shows that our ordering is consistent with the expert ordering.

Manually ordering the images is nontrivial for researchers who are unfamiliar with the developmental progression, and can be tedious and time-consuming for those who are.
%
In contrast, our method requires relatively little computational time; registering and ordering this fixed imaging data set took $15$~seconds on an Intel Core i7 2.93 GHz processor. %for $100$ images of $100 \times 100$ pixels, using 10$^{\circ}$ rotational discretizations.
%
The computational time is a function of the number of images in the data set, the number of pixels in each point, and the angular resolution to compute the pairwise rotations (see \SI). 
%
Empirically, the CPU time required was $\mathcal{O}(n^{1.5})$ in the number of images, and $\mathcal{O}(n)$ in the number of pixels and the angular resolution.
%
Furthermore, the computation of the pairwise rotational alignments, which is the most time-intensive portion of the calculation, is trivially parallelizable.
%
The requisite user intervention and parameter tuning required for our method is relatively minimal.
%
Images must first be preprocessed so that the Euclidean distance between the pixels is informative.
%
Two algorithmic parameters, the angular discretization to compute the pairwise alignments and the diffusion maps kernel scale which determines which data points are ``close'' (see ~\fig~\ref{subfig:dmaps1} and \SI), must also be defined.
%
We found that the results are robust to both of these parameters, and that consistent parameter values served well for all data sets presented here. 
%
%In our examples, we use 10$^{\circ}$ angle discretizations for the pairwise alignments, and one-tenth of the median pairwise distance for the kernel scale. 
% (5$^{\circ}$--30$^{\circ}$ angle discretizations and kernel scales of one-fifth to one-twentieth of the median pairwise distances all yielded acceptable results)
%
The tasks of image preprocessing and parameter selection are relatively simple compared to manual registration and ordering of images, and so this methodology is promising for much larger imaging data sets which are impractical to evaluate manually.

@@

We presented a unified approach to temporal ordering and registration of images in studies of developmental dynamics. 
%
To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other. 
%
In particular, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \cite{yuan2014automated, surkova2008characterization}.  
%
Temporal ordering of large-scale data which did not require registration was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA.
%
Automated temporal ordering of such data sets was accomplished by first projecting the data onto a low-dimensional subspace spanned by the leading principal or independent components, and then solving a traveling salesman problem or constructing a minimum spanning tree on the projected data to order multiple snapshots \cite{anavy2014blind, trapnell2014dynamics, gupta2008extracting}. 
%
The presented approach addresses image registration and ordering in a single step and orders the data without first projecting them on a low-dimensional subspace.
%; vector diffusion maps can also order images when global registration of the entire data set is topologically impossible \cite{zhao2014rotationally}.
%
%The task of image registration has been widely studied \cite{zitova2003image}, for applications such as face recognition \cite{rowley1998rotation}, medical image registration \cite{hajnal2010medical}, and texture classification \cite{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \cite{ian1998statistical} (such as the eyes in face recognition applications \cite{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications. 
%
Although we have presented results for a specific type of data (cross-sectional images taken along the dorsoventral axis of developing {\em Drosophila} embryos), our methodology only relies on computing distances between images and may be readily extended to many other imaging data sets.

@@

Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
This necessitates the development of automated methodologies to organize such large data sets;
such methods should be sufficiently general to address a wide variety of applications
%
Vector diffusion maps allow us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with parameters to describe each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several parameters which concisely and comprehensively describe data set.
%
This parameterization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages. 
%
Images taken from different viewing directions can be analyzed, as the vector diffusion maps parameterization will organize the images according to the viewing angle.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \cite{krzic2012multiview, ichikawa2014live, rubel2010coupling}.  
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
